---
title: "Getting Started with MDSMClust"
author: Xinyue Wang
# output: pdf_document
# vignette: >
#   %\VignetteIndexEntry{MDSMClust}
#   %\VignetteEngine{knitr::knitr}
#   %\VignetteEncoding{UTF-8} 
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{MDSMClust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Install MDSMClust from GitHub  
# Detailed instructions can be found here: https://github.com/wxy929/MDS/blob/main/README.md

library(MDSMClust)
```

# Overview

The MDS for Microbiome Clustering package (v0.1.0) contains functions that conduct classical multidimensional scaling (MDS) with seven different distances or fit Dirichlet Multinomial Models (DMM) to a count matrix first, and then calculate agreement indices between any two partitions for a data set. It also contains four data sets which are used to simulate counts or analyzed as real data sets.


# Examples for the Function MDS

## Data Preparation

Take two real data sets as examples.

```{r}
data("Martinez")
str(Martinez, list.len = 3)

data("Smits")
str(Smits, list.len = 3)
```

## Conduct MDS and Evaluate the Clustering Results

### Martinez

```{r}
data("Martinez")

# get the tree information - it's necessary when you choose a distance which needs the tree information
tree = Martinez$tree

# get the OTU matrix - rows are samples and columns are taxa
otu.table = Martinez$otutable
otu.count = t(otu.table)
dim(otu.count)

# invoke the function MDS - you need to decide the parameters first
table(Martinez$sampleinfo)
MDS(otu.count, distance = "bray", mdsrank = 5, est = "gap", K.max = 16,
group = as.numeric(as.factor(Martinez$sampleinfo)), randMethod = "MA")
```

### Smits

```{r message=FALSE}
data("Smits")

# get the tree information - it's necessary when you choose a distance which needs the tree information
tree = Smits$tree

# get the OTU matrix - rows are samples and columns are taxa
otu.table = as.matrix(Smits$otutable)
otu.count = t(otu.table)
dim(otu.count)

# invoke the function MDS - you need to decide the parameters first
table(Smits$sampleinfo)

# detailed output can be found in the html file
s = MDS(otu.count, distance = "PhILR", tree = tree, mdsrank = 5, est = "si", K.max = 8, group = as.numeric(as.factor(Smits$sampleinfo)), randMethod = "HA", seed = 2021)
s$bestk
s$ARI
table(s$cluster)
```

# Examples for the Function DMM - Martinez

```{r eval=FALSE}
data(Martinez)
tree = Martinez$tree
otu.table = Martinez$otutable
otu.count = t(otu.table)

DMM(otu.count, group = as.numeric(as.factor(Martinez$sampleinfo)), k = 5, trim = TRUE, threshold = 5)
```

# Example Code to Reproduce DM Simulation in the paper

```{r eval=FALSE}
rm(list = ls())

# library(SpiecEasi)
# library(philr)
# library(fpc)
# library(MiSPU)
# library(GUniFrac)
library(dirmult)
# library(cluster)
# library(clues)
library(DirichletMultinomial)
# library(MASS)
# library(vegan)
# library(mvtnorm)
# library(reshape2)
# library(doSNOW)
library(MDSMClust)

# pre_DM_RData ----------------------------------------------
ini.seed = 2020
n = 400
signal.cluster.zero = 0
signal.zero = 0 
readdepth = 4000


data("DM_throat_info")
throat.tree = throat.info$tree
tree.otu.id = throat.tree$tip.label
tree.par.pi = throat.info$par.pi
tree.par.theta = throat.info$par.theta
otu.cluster = throat.info$otu.cluster
m = length(tree.par.pi)        # 281
n1 = n2 = n3 = n4 = n/4        # 100


groups = c(rep(0,n1), rep(1,n2), rep(2,n3), rep(3,n4))
set.seed(ini.seed)


otu.perc.cluster = NULL
for(i in 1:20){
  otu.perc.cluster = c(otu.perc.cluster, sum(tree.par.pi[which(otu.cluster==i)], na.rm = TRUE) )
}



## simulate the difference - change three clusters
# common -> DM_Common.RData
signal.cluster.pos = c(6, 17, 7)                    
signal.pos = 1.4         

tree.par.pi.grp1 = tree.par.pi
tree.par.pi.grp2 = tree.par.pi
tree.par.pi.grp3 = tree.par.pi
tree.par.pi.grp1[which(otu.cluster==signal.cluster.pos[1])] = tree.par.pi.grp1[which(otu.cluster==signal.cluster.pos[1])] * signal.pos * (1/sqrt(otu.perc.cluster[signal.cluster.pos[1]]))
tree.par.pi.grp2[which(otu.cluster==signal.cluster.pos[2])] = tree.par.pi.grp2[which(otu.cluster==signal.cluster.pos[2])] * signal.pos * (1/sqrt(otu.perc.cluster[signal.cluster.pos[2]]))
tree.par.pi.grp3[which(otu.cluster==signal.cluster.pos[3])] = tree.par.pi.grp3[which(otu.cluster==signal.cluster.pos[3])] * signal.pos * (1/sqrt(otu.perc.cluster[signal.cluster.pos[3]]))

# rare  -> DM_Rare.RData
signal.cluster.pos = c(19, 20, 14)
signal.pos = 1.4         

tree.par.pi.grp1 = tree.par.pi
tree.par.pi.grp2 = tree.par.pi
tree.par.pi.grp3 = tree.par.pi
tree.par.pi.grp1[which(otu.cluster==signal.cluster.pos[1])] = tree.par.pi.grp1[which(otu.cluster==signal.cluster.pos[1])] * signal.pos * (1/sqrt(otu.perc.cluster[signal.cluster.pos[1]]))
tree.par.pi.grp2[which(otu.cluster==signal.cluster.pos[2])] = tree.par.pi.grp2[which(otu.cluster==signal.cluster.pos[2])] * signal.pos * (1/sqrt(otu.perc.cluster[signal.cluster.pos[2]]))
tree.par.pi.grp3[which(otu.cluster==signal.cluster.pos[3])] = tree.par.pi.grp3[which(otu.cluster==signal.cluster.pos[3])] * signal.pos * (1/sqrt(otu.perc.cluster[signal.cluster.pos[3]]))

# random -> DM_Random.RData
signal.pos = 1.4
set.seed(ini.seed)
id = sample(1:281, 60)
id1 = which(names(tree.par.pi) %in% tree.otu.id[id[1:20]])
id2 = which(names(tree.par.pi) %in% tree.otu.id[id[21:40]])
id3 = which(names(tree.par.pi) %in% tree.otu.id[id[41:60]])

tree.par.pi.grp1 = tree.par.pi
tree.par.pi.grp2 = tree.par.pi
tree.par.pi.grp3 = tree.par.pi
tree.par.pi.grp1[id1] = tree.par.pi.grp1[id1] * signal.pos * (1/sqrt(sum(tree.par.pi[id1], na.rm = TRUE)))
tree.par.pi.grp2[id2] = tree.par.pi.grp2[id2] * signal.pos * (1/sqrt(sum(tree.par.pi[id2], na.rm = TRUE)))
tree.par.pi.grp3[id3] = tree.par.pi.grp3[id3] * signal.pos * (1/sqrt(sum(tree.par.pi[id3], na.rm = TRUE)))



# Simulation in One of Three Possible Scenarios --------------

# load("DM_Common.RData") from above code
rare = FALSE    # TRUE if consider the rarefication

set.seed(2024)  # seeds from 2024 to 2523 - avg of 500 seeds reported in the paper
# simulate data from Dirichlet-multinomial distribution
otu.countsim = rbind( simPop(J=n1, n=readdepth[1:100],   pi=tree.par.pi, theta=tree.par.theta)$data,
                      simPop(J=n2, n=readdepth[101:200], pi=tree.par.pi.grp1, theta=tree.par.theta)$data,
                      simPop(J=n3, n=readdepth[201:300], pi=tree.par.pi.grp2, theta=tree.par.theta)$data,
                      simPop(J=n4, n=readdepth[301:400], pi=tree.par.pi.grp3, theta=tree.par.theta)$data)

otu.propsim = otu.countsim/rowSums(otu.countsim) 
colnames(otu.propsim) = tree.otu.id
colnames(otu.countsim) = tree.otu.id

# rarefication
if(rare){  
  readdepth.rarefied = min(readdepth)
  otu.countsim.rarefied = matrix(0, ncol = ncol(otu.countsim), nrow = nrow(otu.countsim))
  for(i in 1:n){
    otu.countsim.rarefied[i, ] = t(rmultinom(n=1, size = readdepth.rarefied, prob = otu.propsim[i, ]))
  }
  colnames(otu.countsim.rarefied) = tree.otu.id
  
  otu.propsim = otu.countsim.rarefied/rowSums(otu.countsim.rarefied)
  colnames(otu.propsim) = tree.otu.id
  otu.countsim = otu.countsim.rarefied
  
  print("Rarified")
}

# then take the simulated "otu.count" as the input to the "MDS" function
```
