---
title: "MDSMClust: Multidimensional Scaling for Microbiome Clustering"
author: Guanhua Chen, Xinyue Wang
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MDSMClust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MDSMClust)
```

# Overview

The MDS for Microbiome Clustering package (v0.1.0) contains functions that conduct classical multidimensional scaling (MDS) with seven different distances or fit Dirichlet Multinomial Models (DMM) to a count matrix first, and then calculate agreement indices between any two partitions for a data set. It also contains four data sets which are used to simulate counts or analyzed as real data sets.

# Dependencies

The following packages are required for functions and examples in the MDS for Microbiome Clustering package: SpiecEasi (>= 1.1.0), BiocManager (>= 1.30.10), philr (>= 1.14.0), fpc (>= 2.2.8), MiSPU (>= 1.0), GUniFrac (>= 1.1), miProfile (>= 1.0), dirmult (>= 0.1.3.4), cluster (>= 2.1.0), clues (>= 0.6.2.2), DirichletMultinomial (>= 1.30.0), MASS (>= 7.3.53), vegan (>= 2.5.6), reshape2 (>= 1.4.4). Among these packages, the most special one is *SpiecEasi* because you can only install it by the command  "install_github("zdk123/SpiecEasi")". And *philr* and *DirichletMultinomial* are installed through *BiocManager*. Be careful with the package *miProfile*, it is not available on CRAN now and this is the only one you need to install manually before using our package. Other required packages are available on CRAN.

# Examples for Function MDS

## Data Preparation

Take two real data sets as examples.

```{r}
data("Martinez")
str(Martinez, list.len = 3)
data("Smits")
str(Smits, list.len = 3)
```

## Conduct MDS and Evaluate the Clustering Results

### Martinez

```{r}
data("Martinez")

# get the tree information - it's necessary when you choose a distance which needs the tree information
tree = Martinez$tree

# get the OTU matrix - rows are samples and columns are taxa
otu.table = Martinez$otutable
otu.count = t(otu.table)
dim(otu.count)

# invoke the function MDS - you need to decide the parameters first
table(Martinez$sampleinfo)
MDS(otu.count, distance = "bray", mdsrank = 5, est = "gap", K.max = 16,
group = as.numeric(as.factor(Martinez$sampleinfo)), randMethod = "MA")
```

### Smits

```{r message=FALSE}
data("Smits")

# get the tree information - it's necessary when you choose a distance which needs the tree information
tree = Smits$tree

# get the OTU matrix - rows are samples and columns are taxa
otu.table = as.matrix(Smits$otutable)
otu.count = t(otu.table)
dim(otu.count)

# invoke the function MDS - you need to decide the parameters first
table(Smits$sampleinfo)
s = MDS(otu.count, distance = "PhILR", tree = tree, mdsrank = 5, est = "si", K.max = 8, group = as.numeric(as.factor(Smits$sampleinfo)), randMethod = "HA", seed = 2021)
s$bestk
s$ARI
table(s$cluster)
```

# Examples for Function DMM - Martinez

```{r}
data(Martinez)
tree = Martinez$tree
otu.table = Martinez$otutable
otu.count = t(otu.table)

DMM(otu.count, group = as.numeric(as.factor(Martinez$sampleinfo)), k = 5, trim = T, threshold = 5)
```
